---
title: "HR Exploration"
output: html_notebook
---

Home run data downloaded from [BaseballSavant](https://baseballsavant.mlb.com).

Read all csv files in the `data` directory into a single dataframe.

Check for null values. Also need to strip out inside-the-park home runs, where the batter did not hit the ball out of the field of play.

```{r}

library(dplyr)
library(tidyr)

set.seed(42274)

data_dir <- "./data/"
data_files <- list.files(path=data_dir, pattern="*.csv", full.names=TRUE)
data <- do.call(rbind, lapply(data_files, function (x) read.csv(x, na.strings=c("null", "0", "0.0"))))

a <- as.Date(data$game_date, format="%Y-%m-%d")
b <- as.Date(data$game_date, format="%m/%d/%Y") # Produces NA when format is not "%m/%d/%Y"
a[is.na(a)] <- b[!is.na(b)]
data$date <- a

hr_hits = grep(pattern="inside-the-park", x=data$des, invert=TRUE)
data <- data[hr_hits, ]

```


Let's examine the rate at which home runs are hit in a given season.


```{r}

by_year <- data %>% group_by(game_year)

years <- c(2011:2017)
colors <- rainbow(length(years))
plot(1, type='l', xlab='Day of Season', ylab='Cumulative Home Run Count', xlim=c(0,200),ylim=c(0,6000))
for (y in years) {
  year_data <- by_year %>% filter(game_year == y) %>% group_by(game_date)
  sum <- cumsum(count(year_data)$n)
  lines(sum, col=colors[y-2010])
}
legend("topright", legend = years, col = colors, lwd = 1,
       cex = 0.5)

```

There does appears to be an inflection point around Day 50 of the 2016 season. After that point, the slope of the 2016 and 2017 lines is steeper.

Unfortunately, we only have home run distance information for the 2015 season onward. First, drop all entries with no distance data. Then we can look at the distribution of distances.

```{r}

clean_data <- data %>% drop_na(hit_distance_sc)
boxplot(hit_distance_sc ~ game_year, data=clean_data, xlab='Season', ylab='Home Run Distance')

```

Next we can look at the distributions

```{r}

years <- c(2015:2017)
colors <- rainbow(length(years))
plot(1, type='l', xlab='Home Run Distance', ylab='Number of Hits', xlim=c(275,525),ylim=c(0,115))
for (y in years) {
  year_data <- clean_data[clean_data$game_year == y, ]
  counts <- table(year_data$hit_distance_sc)
  lines(counts, col=colors[y-2014], type="l")
}
legend("topright", legend = years, col = colors, lwd = 1,
       cex = 0.5)

```

What is we segment the data at the inflection point in the 2016 season?

```{r}

clean_data <- clean_data[clean_data$game_year %in% c(2015, 2016, 2017), ]
first_day_2016 <- first(sort(clean_data[clean_data$game_year == 2016, ]$date))

set_ball_type <- function (curr_date, test_date) {
  if (curr_date['date'] <= test_date) {
    "pre"
  }
  else
  {
    "post"
  }
}

ball_type <- apply(clean_data, 1, set_ball_type, first_day_2016 + 50)
clean_data$ball_type = as.factor(ball_type)
clean_data$ball_type <- ordered(clean_data$ball_type, c("pre", "post"))

boxplot(hit_distance_sc ~ ball_type, data=clean_data, xlab='Ball Type', ylab='Home Run Distance')

```

```{r}

ball_types <- c("pre", "post")
colors <- rainbow(length(ball_types))
plot(1, type='l', xlab='Home Run Distance', ylab='Number of Hits', xlim=c(275,525),ylim=c(0,200))
for (bt in ball_types) {
  type_data <- clean_data[clean_data$ball_type == bt, ]
  counts <- table(type_data$hit_distance_sc)
  lines(counts, col=colors[type_data[1, ]$ball_type], type="l")
}
legend("topright", legend = ball_types, col = colors, lwd = 1,
       cex = 0.5)

```
